\documentclass{bioinfo}
\copyrightyear{2019} \pubyear{2019}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Applications Note}

\begin{document}
\firstpage{1}

\subtitle{Genetic and population analysis}

\title[Unified power analysis and forensics of association studies]{U-PASS: unified power analysis and forensics for qualitative traits in genetic association studies}
\author[Gao \textit{et~al}.]{Zheng Gao\,$^{\text{\sfb 1,}*}$, Jonathan Terhorst\,$^{\text{\sfb 1}}$, Cristopher Van Hout\,$^{\text{\sfb 2}}$, and Stilian Stoev\,$^{\text{\sfb 1}}$}
\address{$^{\text{\sf 1}}$Department of Statistics, University of Michigan, Ann Arbor, MI 48109, USA and \\
$^{\text{\sf 2}}$Regeneron Genetics Center, Tarrytown, NY 10591
USA.}

\corresp{$^\ast$To whom correspondence should be addressed.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\abstract{
\textbf{Summary:} 
Despite the availability of existing calculators for statistical power analysis in genetic association studies, there has not been a model-invariant and test-independent tool that allows for both planning of prospective studies and systematic review of reported findings.
In this work, we develop a web-based application U-PASS (Unified Power analysis of ASsociation Studies), implementing a unified framework for the analysis of common association tests for binary qualitative traits.
The application quantifies the shared asymptotic power limits of the common association tests, and visualizes the fundamental statistical trade-off between risk allele frequency (RAF) and odds ratio (OR). 
The application also addresses the applicability of asymptotics-based power calculations in finite samples, and provides guidelines for single-SNP-based association tests. 
In addition to designing prospective studies, U-PASS enables researchers to retrospectively assess the statistical validity of previously reported associations.\\
\textbf{Availability and implementation:} U-PASS is an open-source R Shiny application. A live instance is hosted at {https://power.stat.lsa.umich.edu}. 
Source is available on {https://github.com/Pill-GZ/U-PASS}.\\ %License information will be available in the repo?
\textbf{Contact:} \href{gaozheng@umich.edu}{gaozheng@umich.edu}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}
\maketitle

\section{Introduction}
\label{sec:intro}
\input{1.intro.tex}

\vspace*{-16pt}
\section{Methods and features}
\label{sec:results}
\input{2.results.tex}

\vspace*{-16pt}
\section{Implementation}
\label{sec:implementaion}
\input{3.implementation.tex}

% \vspace*{-10pt}
% \section*{Acknowledgements}
% 
% We thank so and so for their invaluable comments and suggestions on the early versions of the software and this paper.

\vspace*{-16pt}
\section*{Funding}

This work is partially supported by NSF Grant DMS-1830293, ATD.
\vspace*{-16pt}

%\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
%\bibliographystyle{plain}
%
%\bibliography{Document}


\begin{thebibliography}{}

\bibitem[Derkach {\it et~al}., 2017]{Derkach17}
Derkach, A., Zhang, H., \& Chatterjee, N. (2017). 
Power Analysis for Genetic Association Test (PAGEANT) provides insights to challenges for rare variant association studies. 
{\it Bioinformatics}, {\bf 34}, 1506-1513.

% \bibitem[Gonz\'{a}lez {\it et~al}., 2008]{Gonzalez08}
% Gonz\'{a}lez, J. R., Carrasco, J. L., Dudbridge, F., Armengol, L., Estivill, X., \& Moreno, V. (2008). 
% Maximizing association statistics over genetic models. 
% {\it Genetic Epidemiology}, {\bf 32(3)}, 246-254.

\bibitem[Dominguez-Cruz {\it et~al}., 2018]{Dominguez-Cruz18}
Dom\'inguez-Cruz, M. G., de Lourdes Mu\~noz, M., Totomoch-Serra, A., Garc\'ia-Escalante, M. G., Burgue\~no, J., Valadez-Gonz\'alez, N., ... \& D\'iaz-Badillo, \'A. (2018). 
Pilot genome-wide association study identifying novel risk loci for type 2 diabetes in a Maya population. 
{\it Gene}, {\bf 677}, 324-331.

\bibitem[Goodwin {\it et~al}., 2016]{Goodwin16}
Goodwin, S., McPherson, J. D., \& McCombie, W. R. (2016). 
Coming of age: ten years of next-generation sequencing technologies. 
{\it Nat. Reviews Genetics}, {\bf 17}, 333.

\bibitem[Ioannidis, 2005]{Ioannidis05}
Ioannidis, J. P. (2005).
Why most published research findings are false. 
{\it PLoS medicine}, {\bf 2}, e124.

\bibitem[Johnson {\it et~al}., 2017]{Johnson17}
Johnson, J. L., \& Abecasis, G. R. (2017). 
GAS Power Calculator: web-based power calculator for genetic association studies. 
{\it bioRxiv}, 164343.

% \bibitem[Li {\it et~al}., 2008]{Li08}
% Li, Q., Zheng, G., Li, Z., \& Yu, K. (2008). 
% Efficient approximation of P‐value of the maximum of correlated tests, with applications to genome‐wide association studies. 
% {\it Annals of human genetics}, {\bf 72(3)}, 397-406.

\bibitem[MacArthur {\it et~al}., 2016]{MacArthur16}
MacArthur, J., Bowler, E., Cerezo, M., Gil, L., Hall, P., Hastings, E., ... \& Pendlington, Z. M. (2016). 
The new NHGRI-EBI Catalog of published genome-wide association studies (GWAS Catalog). 
{\it Nucleic acids research}, {\bf 45}, D896-D901.

\bibitem[Menashe {\it et~al}., 2008]{Menashe08}
Menashe, I., Rosenberg, P. S., \& Chen, B. E. (2008). 
PGA: power calculator for case-control genetic association analyses. 
{\it BMC genetics}, {\bf 9}, 36.

\bibitem[Purcell {\it et~al}., 2003]{Purcell03}
Purcell, S., Cherny, S. S., \& Sham, P. C. (2003). 
Genetic Power Calculator: design of linkage and association genetic mapping studies of complex traits. 
{\it Bioinformatics}, {\bf 19}, 149-150.

\bibitem[Sham, 1998]{Sham98}
Sham, P. C. (1998). {\it Statistics in human genetics}. Wiley.

\bibitem[Sham  \& Purcell, 2014]{Sham14}
Sham, P. C., \& Purcell, S. M. (2014).
Statistical power and significance testing in large-scale genetic studies. 
{\it Nature Reviews Genetics}, {\bf 15}, 335.

\bibitem[Skol {\it et~al}., 2006]{Skol06}
Skol, A. D., Scott, L. J., Abecasis, G. R., \& Boehnke, M. (2006). 
Joint analysis is more efficient than replication-based analysis for two-stage genome-wide association studies. 
{\it Nature genetics}, {\bf 38}, 209.

\bibitem[Wang {\it et~al}., 2014]{Wang14}
Wang, G. T., Li, B., Lyn Santos-Cortez, R. P., Peng, B., \& Leal, S. M. (2014). 
Power analysis and sample size estimation for sequence-based association studies. 
{\it Bioinformatics}, {\bf 30}, 2377-2378.

% \bibitem[Xiao \& Boehnke, 2009]{Xiao09}
% Xiao, R., \& Boehnke, M. (2009). 
% Quantifying and correcting for the winner's curse in genetic association studies. 
% {\it Genetic Epidemiology}, {\bf 33(5)}, 453-462.

% \bibitem[Lee {\it et~al}., 2014]{Lee2014}
% Lee, S., Abecasis, G. R., Boehnke, M., \& Lin, X. (2014). Rare-variant association analysis: study designs and statistical tests. 
% {\it The American Journal of Human Genetics}, {\bf 95(1)}, 5-23.

\end{thebibliography}
\end{document}

